DATA 621 - Business Analytics and Data Mining

### 2020-12-04 Final_project/Final Project Paper - Group 2.Rmd
Reached a decent stopping point.

---
title: "DATA 621 - Business Analytics and Data Mining"
subtitle: 'Fall 2020 - Group 2 - Final Project Paper'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "12/14/2020"
output:
  pdf_document:
    extra_dependencies:
      amsmath: null
      inputenc: utf8
      xcolor: dvipsnames
      setspace: singlespacing
    toc: FALSE
    toc_depth: 3
urlcolor: purple
bibliography: finalprojectrefs.bib
csl: ieee-with-url.csl
abstract: "Using data from YouTube, we attempt to predict the number of views a video will receive using criteria such as location, category, number of likes, number of dislikes, and number of comments. Using forms of linear regression we have covered this semester, we will test various combinations of features for predictive power. [FILL IN OUTCOMES HERE]\\par\\textbf{Keywords:} Youtube, linear regression, count regression, R"
---

@@

# Introduction
YouTube has changed the future of video entertainment forever [@Moylan2015ADo]. 
In 2019, the platform was estimated to have between \$16 billion and \$25
billion in revenue [@Wakabayashi2019YIa]. YouTube’s model connects a user’s
creativity with a desire for global recognition [@GoogleUtY]. Before YouTube,
international fame was not conceivable outside of a standard television or movie
studio. Today, creators from all over the world are gaining international
prominence using their own equipment and space. “YouTube is central to today’s
video ecosystem,” says Enders Analysis research analyst Jamie McGowan Stuart
[@Foster2020Yt1]. The current top channel “Vlad and Nikita” earns around \$312
thousand per video [@TYL]. According to a survey from Google, 6 out of 10 people
already prefer online video platforms or streaming services over live TV
[@OBTlv]. There are researchers predicting that by 2025---four years now---half
of viewers under the age of 32 will not pay TV service [@McQuivey2015B25].
Understanding some of the underpinnings of what generates views is beneficial to
anyone entering, or already in, the YouTube world.

@@

# Literature Review
With the popularity of Youtube, this is not a new question. Approaches this
question in the greater data science network include those based on more
advanced machine learning techniques such as SGD or neural net classifiers
[@LEZ2019YVP]. Others leveraged NLP and specially engineered features such as
"clickbait" or "NSFW" tags [@Srinivasan2017YVP]. These attempts usually used the
same dataset as we are.

@@

To minimize division by 0 errors, `balance` is defined as follows:
\[
balance =\begin{cases}
\mathrm{ratings\;enabled}\quad\frac{\min(likes, dislikes)}{\max(likes, dislikes)}\\
\mathrm{ratings\;disabled}\quad 1
\end{cases}
\]
This constrains the ratio to the interval \([0, 1]\) with a maximum of 1 when
the two are equal. Now, division by 0 can only occure when both are 0. This 
usually occurs when ratings are disabled. There is only `r ldOn` case out of the
`r nobsUSExt` observations where the ratings were not disabled, yet there are
neither likes nor dislikes. As this is a distinct incongruity for YouTube, we
will remove that one observation from the data. When ratings are disabled,
perforce there is no disparity so the ratio will be set to 1.

@@

```{r dummyVarsAA}
# Create dummy Variables. These will be used for the next three models
modDum <- dummyVars(views ~ likes + dislikes + comments + tags + category +
                   cmtDisabled + rtgDisabled + vidError + disp,
                   data = seenSet, fullRank = TRUE)
seenX <- predict(modDum, seenSet)
seenY <- seenSet$views
```

@@

```{r lm3Table}
# The only way to extract the coefficients from an elastic net model in R
# Is to predict them using the optimal tuning values

@@

Nevertheless, the *relative* magnitude and sign of the parameters are in line
with our expectations. Both `likes` and `dislikes` are positively correlated
with views and `comments` is negatively correlated. A few more categories join
`Film & Animation` as contributing to excess views over the baseline
`Entertainment`. But `Nonprofits & Activism` has the largest factor by far, and
it is negative, which stands in contrast to the second model. Lastly, having a
larger disparity (ratio of likes to dislikes closer to 1of likes tends to reduce views.
Conflict seems to be good for Youtubers!

@@

```{r modTest}
tstDum <- dummyVars(views ~ likes + dislikes + comments + tags + category +
                   cmtDisabled + rtgDisabled + vidError + disp,
                   data = hideSet, fullRank = TRUE)
hideX <- predict(tstDum, hideSet)
hideY <- hideSet$views
lm1P <- predict(lm1, newdata = hideX)
lm2P <- predict(lm2, newdata = hideX)
lm3P <- predict(lm3, newdata = hideX)
compTable <- data.table(Model = c('LM', 'GLM: Gauss+Log', 'ElasticNet'),
                        RMSE = c(RMSE(lm1P, hideY), RMSE(lm2P, hideY),
                                 RMSE(lm3P, hideY)),
                        R2 = c(R2(lm1P, hideY, formula = 'traditional'),
                               R2(lm2P, hideY, formula = 'traditional'),
                               R2(lm3P, hideY, formula = 'traditional')),
                        MAE = c(MAE(lm1P, hideY), MAE(lm2P, hideY),
                                MAE(lm3P, hideY)))
kable(compTable, digits = 3L, format.args = list(big.mark = ','),
      caption = "Model Performance on Test Set")
```

# Discussion & Conclusions
The three models contain both intuitive and counter-intuitive results. As
expected, videos with more ratings tend to have more views. However, as likes
outnumber dislikes, that tends to reduce the number of views. Having a video
witk

# Appendix
## Statistical Appendix
A more detailed analysis of the empirical statistics for the numeric data is 
found in the table below.

\footnotesize
```{r statsTable}
kable(statsN, digits = 2,
      caption = "Statistics for Numeric Variables by Country & Global",
      format.args = list(big.mark = ","))
```
\normalsize

@@

<!-- Statistical Appendix-->
```{r statsTable}
```

<!-- References needs to be last -->  
# References

>>>>

---
title: "DATA 621 - Business Analytics and Data Mining"
subtitle: 'Fall 2020 - Group 2 - Final Project Paper'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "12/11/2020"
output:
  pdf_document:
    extra_dependencies:
      amsmath: null
      inputenc: utf8
      xcolor: dvipsnames
      setspace: singlespacing
    toc: TRUE
    toc_depth: 3
urlcolor: purple
bibliography: finalprojectrefs.bib
csl: ieee-with-url.csl
abstract: "Using data from YouTube, we attempt to predict the number of views a video will receive using criteria such as location, category, number of likes, number of dislikes, and number of comments. Using forms of linear regression we have covered this semester, we will test various combinations of features for predictive power. YouTubers are advised to allow viewers to interact via ratings and to post content which will engender heated discussion. \\par\\textbf{Keywords:} Youtube, linear regression, elastic net, R"
---

@@

# Introduction
YouTube has changed the future of video entertainment forever [@Moylan2015ADo]. 
In 2019, the platform was estimated to have between \$16 billion and \$25
billion in revenue [@Wakabayashi2019YIa]. YouTube’s model connects a user’s
creativity with a desire for global recognition [@GoogleUtY]. Before YouTube,
international fame was not conceivable outside of a standard television or movie
studio. Today, creators from all over the world are gaining international
prominence using their own equipment and space. “YouTube is central to today’s
video ecosystem,” says Enders Analysis research analyst Jamie McGowan Stuart
[@Foster2020Yt1]. The current top channel “Vlad and Nikita” earns around
\$312,000 per video [@TYL]. According to a survey from Google, 6 out of 10
people already prefer online video platforms or streaming services over live TV
[@OBTlv]. There are researchers predicting that by 2025---four years now---half
of viewers under the age of 32 will not pay TV service [@McQuivey2015B25].
Understanding some of the underpinnings of what generates views is beneficial to
anyone entering, or already in, the YouTube world.

@@

# Literature Review
With the popularity of Youtube, this is not a new question. Approaches this
question in the greater data science network include those based on more
advanced machine learning techniques such as SGD or neural net classifiers
[@LEZ2019YVP]. Others leveraged NLP and specially engineered features such as
"clickbait" or "NSFW" tags [@Srinivasan2017YVP]. These attempts often used the
same dataset as we are using.

@@

To minimize division by 0 errors, `balance` is defined as follows:
\[
balance =\begin{cases}
\mathrm{ratings\;enabled}\quad\frac{\min(likes, dislikes)}{\max(likes, dislikes)}\\
\mathrm{ratings\;disabled}\quad 1
\end{cases}
\]
This constrains the ratio to the interval \([0, 1]\) with a maximum of 1 when
the two are equal. Now, division by 0 can only occur when both are 0. This 
usually occurs when ratings are disabled. There is only `r ldOn` case out of the
`r nobsUSExt` observations where the ratings were not disabled, yet there are
neither likes nor dislikes. As this is a distinct incongruity for YouTube, we
will remove that one observation from the data. When ratings are disabled,
perforce there is no disparity so the ratio will be set to 1.

@@

```{r dummyVarsAA}
# Create dummy Variables. These will be used for the next three models
modDum <- dummyVars(views ~ likes + dislikes + comments + tags + category +
                   cmtDisabled + rtgDisabled + vidError + balance,
                   data = seenSet, fullRank = TRUE)
seenX <- predict(modDum, seenSet)
seenY <- seenSet$views
```

@@

```{r lm3Table}
# The only way to extract the coefficients from an elastic net model in R is to
# predict them using the optimal tuning values

@@

Nevertheless, the *relative* magnitude and sign of the parameters are in line
with our expectations. Both `likes` and `dislikes` are positively correlated
with views and `comments` is negatively correlated. A few more categories join
`Film & Animation` as contributing to excess views over the baseline
`Entertainment`. But `Nonprofits & Activism` has the largest factor by far, and
it is negative, which stands in contrast to the second model. Lastly, having a
higher balance---ratio of likes to dislikes closer to 1---tends to increase
views. Conflict seems to be good for YouTubers!

@@

```{r modTest}
# Process the hidden set
tstDum <- dummyVars(views ~ likes + dislikes + comments + tags + category +
                   cmtDisabled + rtgDisabled + vidError + balance,
                   data = hideSet, fullRank = TRUE)
hideX <- predict(tstDum, hideSet)
hideY <- hideSet$views

# Predict using the models
lm1P <- predict(lm1, newdata = hideX)
lm2P <- predict(lm2, newdata = hideX)
lm3P <- predict(lm3, newdata = hideX)

# Compare the results
compTable <- data.table(Model = c('LM', 'GLM: Gauss+Log', 'ElasticNet'),
                        RMSE = c(RMSE(lm1P, hideY), RMSE(lm2P, hideY),
                                 RMSE(lm3P, hideY)),
                        R2 = c(R2(lm1P, hideY, formula = 'traditional'),
                               R2(lm2P, hideY, formula = 'traditional'),
                               R2(lm3P, hideY, formula = 'traditional')),
                        MAE = c(MAE(lm1P, hideY), MAE(lm2P, hideY),
                                MAE(lm3P, hideY)))
kable(compTable, digits = 3L, format.args = list(big.mark = ','),
      caption = "Model Performance on Test Set")
```

# Discussion & Conclusions
The three models contain both intuitive and counter-intuitive results. As
expected, videos with more ratings tend to have more views. However, as likes
outnumber dislikes, that tends to reduce the number of views. All the models
agree that having a video whose likes and dislikes are close in magnitude
increases the propensity for views.

The elasticNet model performed best, and it's clear that the log link is
inferior to the identity link for this data set. If there is any clear takeaways
for prospective YouTubers, it would be:

  * Make sure your videos will engender heated discussions. You want as many
  *dislikes* as likes. That seems to drive attention to your video!
  * Keep your ratings enabled. Allowing users to rate your video drives
  attention to them!
  * Perhaps disable comments, though.
  * Stay **away** from activism. 
  * Talk about films or comics. This probably includes the anime and manga
  phenomena.
  * Having more tags is better than having fewer.

The main limitation of this analysis is that the linear model, even with
penalization, is probably not the best model for human behavior. There are
likely non-linear effects and tipping points which could be much better captured
by tree-based models or even SVMs with the appropriate kernel.

Also, this analysis may be valuable for someone planning a video, but it does
not help them react once the video is posted, as we did not investigate any
changes over time.

Lastly, we encountered problems trying to aggregate disjoint data by country
without falling prey to gross overcounting.

Areas of future work would include investigating more sophisticated algorithms,
analyzing not only final views but growth patterns, and either finding better
sources or methods to increase the available data to include other countries
without overcounting.

\clearpage
# Appendix
## Statistical Appendix
### Table of Numeric Predictors
A more detailed analysis of the empirical statistics for the numeric data is 
found in the table below.

\footnotesize
```{r statsTable}
# Empirical stats Table
kable(statsN, digits = 2, caption = "Table of Numeric Predictors",
      format.args = list(big.mark = ","))
```
\normalsize

### Variable Importance: Model 1

```{r lm1VI}
# Model 1: VarImp 
ggplot(varImp(lm1)) + ggtitle('Variable Importance: Model 1')
```

### Variable Importance: Model 2

```{r lm2VI}
# Model 2: VarImp 
ggplot(varImp(lm2)) + ggtitle('Variable Importance: Model 2')
```

### Variable Importance: Model 3
```{r lm3VI}
# Model 3: VarImp 
ggplot(varImp(lm3)) + ggtitle('Variable Importance: Model 3')
```

### Hyperparameter Tuning: Model 3

```{r lm3Plot}
# Model 3: Tuning paths
ggplot(lm3) + ggtitle('Tuning Parameters: Model 3')
```

@@

<!-- Statistical Appendix-->
```{r statsTable}
```
```{r lm1VI}
```
```{r lm2VI}
```
```{r lm2VI}
```
```{r lm3Plot}
```

<!-- References needs to be last -->
\clearpage
# References
